---
title: 'Data Analysis Assignment #1 (50 points total)'
author: "Jue, Scott"
output:
  html_document: default
---

```{r setup, include = FALSE}
# DO NOT ADD OR REVISE CODE HERE
knitr::opts_chunk$set(echo = FALSE, eval = TRUE)

```


Submit both the .Rmd and .html files for grading. You may remove the instructions and example problem above, but do not remove the YAML metadata block or the first, "setup" code chunk.  Address the steps that appear below and answer all the questions. Be sure to address each question with code and comments as needed.  You may use either base R functions or ggplot2 for the visualizations.

-----

The following code chunk will:

(a) load the "ggplot2", "gridExtra" and "knitr" packages, assuming each has been installed on your machine,
(b) read-in the abalones dataset, defining a new data frame, "mydata,"
(c) return the structure of that data frame, and
(d) calculate new variables, VOLUME and RATIO.

Do not include package installation code in this document. Packages should be installed via the Console or 'Packages' tab. You will also need to download the abalones.csv from the course site to a known location on your machine. Unless a *file.path()* is specified, R will look to directory where this .Rmd is stored when knitting.

```{r analysis_setup1, message = FALSE, warning = FALSE}

# a) Load the ggplot2 and gridExtra packages.

library(ggplot2)
library(gridExtra)
library(knitr)

# b) Use read.csv() to read the abalones.csv into R, assigning the data frame to "mydata."

mydata <- read.csv("abalones.csv", sep = ",", stringsAsFactors = TRUE)


# c) Use the str() function to verify the structure of "mydata." You should have 1036 observations
# of eight variables.

str(mydata)

# d) Define two new variables, VOLUME and RATIO. Use the following statements to define VOLUME and
# RATIO as variables appended to the data frame "mydata."

mydata$VOLUME <- mydata$LENGTH * mydata$DIAM * mydata$HEIGHT
mydata$RATIO <- mydata$SHUCK / mydata$VOLUME

```

-----

### Test Items starts from here - There are 6 sections - Total 50 points

***##### Section 1: (6 points) Summarizing the data.***

(1)(a) (1 point) Use *summary()* to obtain and present descriptive statistics from mydata.  Use *table()* to present a frequency table using CLASS and RINGS. There should be 115 cells in the table you present.  

```{r Part_1a}
summary(mydata)
table(mydata$CLASS, mydata$RINGS)

```

**Question (1 point):  Briefly discuss the variable types and distributional implications such as potential skewness and outliers.**

***Answer: The data set contains both categorical data types such as sex and class and numerical variable types for physical measurements such as legnth, diameter, height, whole weight, shuck weight, number of rings, volume and ratio of shuck weight to volume. The number of sex types (Female, Male, and Infants) is evenly distributed for the data set. The distribution of infant abalones are centered around A2 class while the female and male abalones are centered around the A3 class. When comparing the max values to their respective distributions, we can see that there are some outliers in the whole weight, shuck weight, and volume variables that could lead to potential skewness. ***

(1)(b) (1 point) Generate a table of counts using SEX and CLASS. Add margins to this table (Hint: There should be 15 cells in this table plus the marginal totals. Apply *table()* first, then pass the table object to *addmargins()* (Kabacoff Section 7.2 pages 144-147)).  Lastly, present a barplot of these data; ignoring the marginal totals.  

```{r Part_1b}

s_c <- table(mydata$SEX, mydata$CLASS)
addmargins(s_c)

ggplot(as.data.frame(s_c), aes(Var2, Freq, fill=Var1, beside=FALSE)) + 
  geom_bar(stat="identity", position='dodge') +
  xlab("Class")+ ylab("Frequency") +
  guides(fill=guide_legend(title="SEX")) +
  ggtitle("Class membership, Sex-differentiated") +
  theme_minimal() +
  theme(plot.title = element_text(hjust = 0.5))
  


```

**Essay Question (2 points):  Discuss the sex distribution of abalones.  What stands out about the distribution of abalones by CLASS?**

***Answer: The number of sex types (Female, Male, and Infants) is evenly distributed for the data set. The distribution of infant abalones are centered around A2 class while the female and male abalones are centered around the A3 class. Each class also has more male abalones than female abalones. The male and female distribution has a slight negative skew while the infant distribution is positively skewed. This is expected since infants should make up the majority of the lower classes (A1 and A2) and the adults should make up the majority of the higher classes (A4 and A5). ***

(1)(c) (1 point) Select a simple random sample of 200 observations from "mydata" and identify this sample as "work." Use *set.seed(123)* prior to drawing this sample. Do not change the number 123. Note that *sample()* "takes a sample of the specified size from the elements of x." We cannot sample directly from "mydata." Instead, we need to sample from the integers, 1 to 1036, representing the rows of "mydata." Then, select those rows from the data frame (Kabacoff Section 4.10.5 page 87). 

Using "work", construct a scatterplot matrix of variables 2-6 with *plot(work[, 2:6])* (these are the continuous variables excluding VOLUME and RATIO). The sample "work" will not be used in the remainder of the assignment. 

```{r Part_1c}

set.seed(123)
work <- mydata[sample(x = 1:nrow(mydata), size = 200, replace = FALSE), ]

plot(work[, 2:6])

```

-----

***##### Section 2: (5 points) Summarizing the data using graphics.***

(2)(a) (1 point) Use "mydata" to plot WHOLE versus VOLUME. Color code data points by CLASS.

```{r Part_2a}

ggplot(data = mydata, aes(x = WHOLE, y = VOLUME, col = CLASS)) + geom_point() +
  theme_minimal() + ggtitle("Whole weight, in grams | Volume,"~cm^3) +
    theme(plot.title = element_text(hjust = 0.5))
  

```


(2)(b) (2 points) Use "mydata" to plot SHUCK versus WHOLE with WHOLE on the horizontal axis. Color code data points by CLASS. As an aid to interpretation, determine the maximum value of the ratio of SHUCK to WHOLE.  Add to the chart a straight line with zero intercept using this maximum value as the slope of the line. If you are using the 'base R' *plot()* function, you may use *abline()* to add this line to the plot. Use *help(abline)* in R to determine the coding for the slope and intercept arguments in the functions. If you are using ggplot2 for visualizations, *geom_abline()* should be used.

```{r Part_2b}


ggplot(data = mydata, aes(x = WHOLE, y = SHUCK, col = CLASS)) + geom_point() +
  geom_abline(intercept = 0, slope = max(mydata$SHUCK/mydata$WHOLE)) +
  theme_minimal() + ggtitle("Shuck weight, in grams | Whole weight, in grams") +
  theme(plot.title = element_text(hjust = 0.5))

```

**Essay Question (2 points):  How does the variability in this plot differ from the plot in (a)?  Compare the two displays.  Keep in mind that SHUCK is a part of WHOLE.  Consider the location of the different age classes.**

***Answer: The variability of this plot is less than the variability observed from the plot in (a). Both plots show a positive correlation. In plot (a) volume increases and whole weight increase and in plot (b) shuck weight increases as whole weight increases. When comparing the location of the different age classes, the plots show the older age class (A4 and A5) are larger in volume; however, have less shuck weight with respect to whole weight. We can see in plot (a) the A4 and A5 points start to drift above the other points as whole weight increases. But In plot (b) the A4 and A5 points start to fall below the other points and get farther away from the trend line. This tells us that as the abalones get older they get bigger and heavier, but the shell begins to make up more of the weight than the meat does. ***

-----

***### Section 3: (8 points) Getting insights about the data using graphs.***

(3)(a) (2 points) Use "mydata" to create a multi-figured plot with histograms, boxplots and Q-Q plots of RATIO differentiated by sex. This can be done using *par(mfrow = c(3,3))* and base R or *grid.arrange()* and ggplot2. The first row would show the histograms, the second row the boxplots and the third row the Q-Q plots. Be sure these displays are legible.  

```{r Part_3a}

par(mfrow = c(3,3))
hist(mydata$RATIO[mydata$SEX=="I"], col = "red", main = "Infant Ratio", xlab = "Ratio")
hist(mydata$RATIO[mydata$SEX=="M"], col = "blue", main = "Male Ratio", xlab = "Ratio")
hist(mydata$RATIO[mydata$SEX=="F"], col = "green", main = "Female Ratio", xlab = "Ratio")
boxplot(mydata$RATIO[mydata$SEX=="I"], col = "red", main = "Infant Ratio")
boxplot(mydata$RATIO[mydata$SEX=="M"], col = "blue", main = "Male Ratio")
boxplot(mydata$RATIO[mydata$SEX=="F"], col = "green", main = "Female Ratio")
qqnorm(mydata$RATIO[mydata$SEX=="I"], pch=17, col = "red", main = "QQ Plot for Infant Ratio")
qqline(mydata$RATIO[mydata$SEX=="I"], col = "maroon3")
qqnorm(mydata$RATIO[mydata$SEX=="M"], pch=17, col = "blue", main = "QQ Plot for Male Ratio")
qqline(mydata$RATIO[mydata$SEX=="M"], col = "maroon3")
qqnorm(mydata$RATIO[mydata$SEX=="F"], pch=17, col = "green", main = "QQ Plot for Female Ratio")
qqline(mydata$RATIO[mydata$SEX=="F"], col = "maroon3")


```

**Essay Question (2 points): Compare the displays.  How do the distributions compare to normality? Take into account the criteria discussed in the sync sessions to evaluate non-normality.**

***Answer: The distribution for female ratios appear to be normally distributed and the distributions for male and infants have a slight positive skew. The box plots show that there are outliers for all sexes which likely explains the skewness observed in the histograms.The outliers can also be seen in the QQ plots in the top right corner of the plots for each sex where the points deviate significantly from the line. ***

(3)(b) (2 points) The boxplots in (3)(a) indicate that there are outlying RATIOs for each sex. *boxplot.stats()* can be used to identify outlying values of a vector. Present the abalones with these outlying RATIO values along with their associated variables in "mydata". Display the observations by passing a data frame to the kable() function. Basically, we want to output those rows of "mydata" with an outlying RATIO, but we want to determine outliers looking separately at infants, females and males.

```{r Part_3b}

inf_out_ratios <- boxplot.stats(mydata[mydata$SEX == "I", "RATIO"])$out
m_out_ratios <- boxplot.stats(mydata[mydata$SEX == "M", "RATIO"])$out
f_out_ratios <- boxplot.stats(mydata[mydata$SEX == "F", "RATIO"])$out

inf_or_df <- mydata[mydata$SEX == "I" & mydata$RATIO %in% inf_out_ratios, ]
m_or_df <- mydata[mydata$SEX == "M" & mydata$RATIO %in% m_out_ratios, ]
f_or_df <- mydata[mydata$SEX == "F" & mydata$RATIO %in% f_out_ratios, ]

all_or_df <- rbind(inf_or_df, f_or_df, m_or_df)

kable(all_or_df)
```

**Essay Question (2 points):  What are your observations regarding the results in (3)(b)?**

***Answer: There are more outliers than belong to the infants than adult abalones. Additionally, most of the infant outliers were from the A1 class and most of the adult outliers came from the A3 class. There are a few extreme outliers seen in the infant and female ratios. ***

-----

***### Section 4: (8 points) Getting insights about possible predictors.***

(4)(a) (3 points) With "mydata," display side-by-side boxplots for VOLUME and WHOLE, each differentiated by CLASS There should be five boxes for VOLUME and five for WHOLE. Also, display side-by-side scatterplots:  VOLUME and WHOLE versus RINGS. Present these four figures in one graphic:  the boxplots in one row and the scatterplots in a second row. Base R or ggplot2 may be used.


```{r Part_4a}

grid.arrange(
  ggplot(data = mydata, aes(x = CLASS, y = VOLUME, group = CLASS)) + geom_boxplot(fill = "orange2"),
  ggplot(data = mydata, aes(x = CLASS, y = WHOLE, group = CLASS)) + geom_boxplot(fill = "mediumorchid"),
  ggplot(data = mydata,aes(x = RINGS, y = VOLUME)) + geom_point(),
  ggplot(data = mydata,aes(x = RINGS, y = WHOLE)) + geom_point()
)
```

**Essay Question (5 points) How well do you think these variables would perform as predictors of age?  Explain.**

***Answer: For age classes A1-A3 the volume and whole weight variables would perform well as predictors. However since the box plots show a similar box and whisker for the A4 and A5 class, it would be hard to differentiate between the two classes using these variables. The scatterplots further proves observation as the data points become more spread out (increase variability) as the number of rings increase for both volume and whole weight. Therefore, these variables would not perform well for the older age classes.***

-----

***### Section 5: (12 points) Getting insights regarding different groups in the data.***

(5)(a) (2 points) Use *aggregate()* with "mydata" to compute the mean values of VOLUME, SHUCK and RATIO for each combination of SEX and CLASS. Then, using *matrix()*, create matrices of the mean values. Using the "dimnames" argument within *matrix()* or the *rownames()* and *colnames()* functions on the matrices, label the rows by SEX and columns by CLASS. Present the three matrices (Kabacoff Section 5.6.2, p. 110-111). The *kable()* function is useful for this purpose.  You do not need to be concerned with the number of digits presented.

```{r Part_5a}

mean_vols <- aggregate(VOLUME ~ SEX + CLASS, data = mydata, FUN = mean)
mean_shuck <- aggregate(SHUCK ~ SEX + CLASS, data = mydata, FUN = mean)
mean_ratio <- aggregate(RATIO ~ SEX + CLASS, data = mydata, FUN = mean)

mean_v <- round(matrix(data = mean_vols$VOLUME, nrow = 3, ncol = 5, byrow = FALSE,
       dimnames = list(SEX = c("Female", "Infant", "Male"),
                       CLASS = levels(mydata$CLASS))),2)

mean_s <- round(matrix(data = mean_shuck$SHUCK, nrow = 3, ncol = 5, byrow = FALSE,
       dimnames = list(SEX = c("Female", "Infant", "Male"),
                       CLASS = levels(mydata$CLASS))),2)

mean_r <- round(matrix(data = mean_ratio$RATIO, nrow = 3, ncol = 5, byrow = FALSE,
       dimnames = list(SEX = c("Female", "Infant", "Male"),
                       CLASS = levels(mydata$CLASS))),4)

kable(mean_v, caption = "Mean Volume")
kable(mean_s, caption = "Mean Shuck")
kable(mean_r, caption = "Mean Ratio")

```

(5)(b) (3 points) Present three graphs. Each graph should include three lines, one for each sex. The first should show mean RATIO versus CLASS; the second, mean VOLUME versus CLASS; the third, mean SHUCK versus CLASS. This may be done with the 'base R' *interaction.plot()* function or with ggplot2 using *grid.arrange()*.

```{r Part_5b, fig.width = 9}
grid.arrange(
  ggplot(data = mean_ratio, aes(x = CLASS, y = RATIO, group = SEX, color = SEX)) + geom_line() + geom_point() +
    scale_color_manual(name="SEX", labels=c("Female","Infant","Male"), values=c("red","green","blue")) +
    ggtitle("Mean Ratio per Class") + theme(plot.title = element_text(size = 10, hjust = 0.5)),
  ggplot(data = mean_vols, aes(x = CLASS, y = VOLUME, group = SEX, color = SEX)) + geom_line() + geom_point() +
    scale_color_manual(name="SEX", labels=c("Female","Infant","Male"), values=c("red","green","blue")) +
    ggtitle("Mean Vomlume per Class") + theme(plot.title = element_text(size = 10, hjust = 0.5)),
  ggplot(data = mean_shuck, aes(x = CLASS, y = SHUCK, group = SEX, color = SEX)) + geom_line() + geom_point() +
    scale_color_manual(name="SEX", labels=c("Female","Infant","Male"), values=c("red","green","blue")) +
    ggtitle("Mean Shuck per Class") + theme(plot.title = element_text(size = 10, hjust = 0.5))
)

```

**Essay Question (2 points):  What questions do these plots raise?  Consider aging and sex differences.**

***Answer: The first plot raises the question why does the ratio start to decrease after class A2. Also after A1 the infants have a lower mean ratio for classes A2-A4 than the adults. The second and third plot raise a question as to why females have a higher volume and shuck weight among all age classes. Another question is why volume appears to remain flat between A4 and A5. Similarly, the third plot raises the question why the mean shuck weight increases as the abalone ages but declines after A4. This could suggest that the shuck weight maxes out at A4, then starts to decrease after. ***

5(c) (3 points) Present four boxplots using *par(mfrow = c(2, 2)* or *grid.arrange()*. The first line should show VOLUME by RINGS for the infants and, separately, for the adult; factor levels "M" and "F," combined. The second line should show WHOLE by RINGS for the infants and, separately, for the adults. Since the data are sparse beyond 15 rings, limit the displays to less than 16 rings. One way to accomplish this is to generate a new data set using subset() to select RINGS < 16.  Use ylim = c(0, 1100) for VOLUME and ylim = c(0, 400) for WHOLE.  If you wish to reorder the displays for presentation purposes or use ggplot2 go ahead.

```{r Part_5c}

inf_rings <- mydata[mydata$SEX == "I" & mydata$RINGS < 16, ]
adult_rings <- mydata[(mydata$SEX != "I") & mydata$RINGS < 16, ]

grid.arrange(
  ggplot(data = inf_rings, aes(x = RINGS, y = VOLUME, group = RINGS)) + 
    geom_boxplot(fill = "steelblue2", outlier.colour = "red", outlier.shape =  19, show.legend = FALSE) +
    stat_boxplot(geom ='errorbar') + ggtitle("Infant Volume | Rings") +  
    theme(plot.title = element_text(size = 10, hjust = 0.5)) + ylim(0, 1000),
   ggplot(data = adult_rings, aes(x = RINGS, y = VOLUME, group = RINGS)) + 
    geom_boxplot(fill = "seagreen2", outlier.colour = "red", outlier.shape =  19, show.legend = FALSE) +
    stat_boxplot(geom ='errorbar') + ggtitle("Adult Volume | Rings") +  
    theme(plot.title = element_text(size = 10, hjust = 0.5)),
  ggplot(data = inf_rings, aes(x = RINGS, y = WHOLE, group = RINGS)) + 
    geom_boxplot(fill = "steelblue2", outlier.colour = "red", outlier.shape =  19, show.legend = FALSE) +
    stat_boxplot(geom ='errorbar') + ggtitle("Infant Whole Weight | Rings") +  
    theme(plot.title = element_text(size = 10, hjust = 0.5)) + ylim(0, 400),
  ggplot(data = adult_rings, aes(x = RINGS, y = WHOLE, group = RINGS)) + 
    geom_boxplot(fill = "seagreen2", outlier.colour = "red", outlier.shape =  19, show.legend = FALSE) +
    stat_boxplot(geom ='errorbar') + ggtitle("Adult Whole Weight | Rings") +  
    theme(plot.title = element_text(size = 10, hjust = 0.5)) + ylim(0, 400),
  nrow=2
  )

```

**Essay Question (2 points):  What do these displays suggest about abalone growth?  Also, compare the infant and adult displays.  What differences stand out?**

***Answer: The displays suggest that both infant and adult abalones do the majority of their growing based on volume and whole weight between 5 and 12 rings. After 12 rings the volume and whole weight begin to plateau. When comparing the infant and adult displays, the box plots for the adults show a higher degree of variability as the whisker length and box sizes have a wider range than the whiskers and boxes from the infant box plots. The infant displays also have more outliers than the adult displays as denoted by the red points. ***

-----

***### Section 6: (11 points) Conclusions from the Exploratory Data Analysis (EDA).***


**Conclusions**


**Essay Question 1) (5 points)	Based solely on these data, what are plausible statistical reasons that explain the failure of the original study? Consider to what extent physical measurements may be used for age prediction.**

***Answer: The data makes it difficult to fully utilize physical measurement variables as a way to predict age because the prescence of outliers is creating skewness. It would be beneficial to remove the outliers from the data set to remove the skewness resulting in a normal distribution. Some phyical measurements may used to some extent for age prediction such as volume and whole weight for predicting age classes A1-A3 (as seen in part 4). But predicting age classes A4 and A5 is more difficult due to have more variability. ***

**Essay Question 2) (3 points) Do not refer to the abalone data or study.  If you were presented with an overall histogram and summary statistics from a sample of some population or phenomenon and no other information, what questions might you ask before accepting them as representative of the sampled population or phenomenon?**

***Answer: I would first ask if the sample size is large enough to be a good representation of the total population being sampled. Additionally, I would ask how the sample was obtained, what kind of sampling method was used and if there is there any sampling bias. Next I would ask if there are any outliers in data set and if there is any skewness observed. Lastly, I would ask if there was any null data and if so how they were handled. ***

**Essay Question 3)  (3 points)	Do not refer to the abalone data or study.  What do you see as difficulties analyzing data derived from observational studies? Can causality be determined?  What might be learned from such studies?**

***Answer: Analyzing data derived from observational studies can have several issues. First, outliers could present themselves in the observations and create skewness in the data. There could also be difficulties in achieving the appropriate sample size. There could also be issues with sampling bias and achieving a randomized sample as well. Without a sufficient randomize sample size and a control group causality cannot be determined from observational data. Studies using observational data can provide us with insights about the relationship of the variables such as correlation. ***
