---
title: "Data Analysis Assignment #2 (75 points total)"
author: "Jue, Scott"
output:
  html_document: default
---

```{r setup, include = FALSE}
# DO NOT ADD OR REVISE CODE HERE
knitr::opts_chunk$set(echo = FALSE, eval = TRUE)

```


##Data Analysis #2

```{r analysis_setup1, message = FALSE, warning = FALSE}

# Perform the following steps to start the assignment.
 
# 1) Load/attach the following packages via library():  flux, ggplot2, gridExtra, moments, rockchalk, car.
# NOTE:  packages must be installed via install.packages() before they can be loaded.

library(dplyr)
library(flux)
library(ggplot2)
library(gridExtra)
library(knitr)
library(rockchalk)
library(tidyverse)

# 2) Use the "mydata.csv" file from Assignment #1 or use the file posted on the course site.  Reading
# the files into R will require sep = "" or sep = " " to format data properly.  Use str() to check file
# structure.

mydata <- read.csv("mydata.csv", sep = ",", stringsAsFactors = TRUE)
# mydata <- read.csv(file.path("c:...", "mydata.csv"), sep = ",")
# mydata <- read.csv(file.path("c:/Rabalone/", "mydata.csv"), sep = ",")

str(mydata)

```

### Test Items starts from here - There are 10 sections - total of 75 points ##############

***#### Section 1: (5 points) ####***

(1)(a) Form a histogram and QQ plot using RATIO. Calculate skewness and kurtosis using 'rockchalk.' Be aware that with 'rockchalk', the kurtosis value has 3.0 subtracted from it which differs from the 'moments' package. 

```{r Part_1a, fig.width = 12}
par(mfrow = c(1, 2))
hist(mydata$RATIO)
qqnorm(mydata$RATIO, pch=17, col = "blue", main = "QQ Plot for Ratio")
qqline(mydata$RATIO)

skewness(x = mydata$RATIO)
kurtosis(x = mydata$RATIO)

```

(1)(b) Tranform RATIO using *log10()* to create L_RATIO (Kabacoff Section 8.5.2, p. 199-200). Form a histogram and QQ plot using L_RATIO. Calculate the skewness and kurtosis. Create a boxplot of L_RATIO differentiated by CLASS.

```{r Part_1b, fig.width = 12, fig.height = 8}


mydata$L_RATIO <- log10(mydata$RATIO)
par(mfrow = c(1, 2))
hist(mydata$L_RATIO)
qqnorm(mydata$L_RATIO, pch=17, col = "blue", main = "QQ Plot for Ratio")
qqline(mydata$L_RATIO)

skewness(x = mydata$L_RATIO)
kurtosis(x = mydata$L_RATIO)

ggplot(data = mydata, aes(x = CLASS, y = L_RATIO, group = CLASS)) + geom_boxplot(fill = "orange2") +
  ggtitle("Boxplots of L_RATIO by CLASS") + theme(plot.title = element_text(size = 10, hjust = 0.5))

```

(1)(c) Test the homogeneity of variance across classes using *bartlett.test()* (Kabacoff Section 9.2.2, p. 222). 

```{r Part_1c}
bartlett.test(x = mydata$L_RATIO, g = mydata$CLASS)

```

**Essay Question: Based on steps 1.a, 1.b and 1.c, which variable RATIO or L_RATIO exhibits better conformance to a normal distribution with homogeneous variances across age classes?  Why?** 

***Answer: The L_RATIO variable that was created above in 1.b exhibits better conformance to a normal distribution with similar variances across age classes. We can see in the output of 1.b that the skewness is closer to 0 and that the kurtosis is less than the kurtosis calculated in 1.a for RATIO. Additionally, the p-value from the Bartlett test indicates that the variances of the L_RATIO are relatively equal across all age classes.***


***#### Section 2 (10 points) ####***

(2)(a) Perform an analysis of variance with *aov()* on L_RATIO using CLASS and SEX as the independent variables (Kabacoff chapter 9, p. 212-229). Assume equal variances. Perform two analyses. First, fit a model with the interaction term CLASS:SEX. Then, fit a model without CLASS:SEX. Use *summary()* to obtain the analysis of variance tables (Kabacoff chapter 9, p. 227).

```{r Part_2a}
wo_int_mdl <- aov(L_RATIO ~ CLASS + SEX, data = mydata)
w_int_mdl <- aov(L_RATIO ~ CLASS*SEX, data = mydata)

summary(wo_int_mdl)
summary(w_int_mdl)

```

**Essay Question:  Compare the two analyses.  What does the non-significant interaction term suggest about the relationship between L_RATIO and the factors CLASS and SEX?**

***Answer: The high p-value of interaction term CLASS:SEX suggests that it is not significant in explaining the any of the difference in L_RATIO and that it is not relevant to our model. However, both CLASS and SEX individually are significant based on the low p-values observed in the summary of analysis of variance. ***

(2)(b) For the model without CLASS:SEX (i.e. an interaction term), obtain multiple comparisons with the *TukeyHSD()* function. Interpret the results at the 95% confidence level (*TukeyHSD()* will adjust for unequal sample sizes). 

```{r Part_2b}
TukeyHSD(wo_int_mdl)

```

**Additional Essay Question:  first, interpret the trend in coefficients across age classes. What is this indicating about L_RATIO?  Second, do these results suggest male and female abalones can be combined into a single category labeled as 'adults?' If not, why not?**

***Answer: Due to the high p-value of 0.69, there is no significant difference between the L_RATIO of the A1 and A2 classes. However, all other age class comparisons resulted in a low p-value (less than .05), so we can conclude that the differences are statistically significant. This indicates that the L_RATIO would make a good predictor of class outside of the A1 and A2 comparison. The results of the SEX comparison suggest that male and female abalones can be combined into a single "adult" category. This is because the M-F comparison results in a very high p-value of .94 which suggests that there is no significant difference between the two different sexes. However, both F-I and M-I comparisons result in low p-values. This suggests that if we combined F and M into an "adult" category, then the "adult" class would result in a significant difference when compared to the Infant class.***


***####  Section 3: (10 points) ####***

(3)(a1) Here, we will combine "M" and "F" into a new level, "ADULT". The code for doing this is given to you. For (3)(a1), all you need to do is execute the code as given.

```{r Part_3a1}
# Here, we show how to define the new variable TYPE using only base R functions:

mydata$TYPE <- factor(ifelse(mydata$SEX == "I", "I", "ADULT"))
table(mydata$TYPE)


```

(3)(a2)  Present side-by-side histograms of VOLUME. One should display infant volumes and, the other, adult volumes. 

```{r Part_3a2, fig.width = 12}
par(mfrow = c(1, 2))
hist(x = mydata$VOLUME[mydata$TYPE == "I"], xlab = "Volume", main = "Histogram of Infant Volume")
hist(x = mydata$VOLUME[mydata$TYPE == "ADULT"], xlab = "Volume", main = "Histogram of Adult Volume")
```


**Essay Question: Compare the histograms.  How do the distributions differ? Are there going to be any difficulties separating infants from adults based on VOLUME?**

***Answer: The infant volume histogram is right skewed while the adult volume histogram results in a distribution that is closer to normal. There will be difficulties in separating infants from adults based on volume because, there is significant overlap in distributions between the infant group and adult group between volumes 0-250.***

(3)(b) Create a scatterplot of SHUCK versus VOLUME and a scatterplot of their base ten logarithms, labeling the variables as L_SHUCK and L_VOLUME. Please be aware the variables, L_SHUCK and L_VOLUME, present the data as orders of magnitude (i.e. VOLUME = 100 = 10^2 becomes L_VOLUME = 2). Use color to differentiate CLASS in the plots. Repeat using color to differentiate by TYPE. 

```{r Part_3b, fig.width = 12, fig.height = 8}

mydata$L_SHUCK <- log10(mydata$SHUCK)
mydata$L_VOLUME <- log10(mydata$VOLUME)

grid.arrange(
ggplot(data = mydata, aes(x = VOLUME, y = SHUCK,  col = CLASS)) + geom_point() + theme(legend.position="bottom"),
ggplot(data = mydata, aes(x = L_VOLUME, y = L_SHUCK,  col = CLASS)) + geom_point() + theme(legend.position="bottom"),
ggplot(data = mydata, aes(x = VOLUME, y = SHUCK,  col = TYPE)) + geom_point() + theme(legend.position="bottom"),
ggplot(data = mydata, aes(x = L_VOLUME, y = L_SHUCK,  col = TYPE)) + geom_point() + theme(legend.position="bottom")
)
```

**Additional Essay Question:  Compare the two scatterplots. What effect(s) does log-transformation appear to have on the variability present in the plot?  What are the implications for linear regression analysis? Where do the various CLASS levels appear in the plots? Where do the levels of TYPE appear in the plots?**

***Answer: Performing log-transformations on the SHUCKE and VOLUME variables decrease the variability. The scatterplots without log-transformation are more spread out and the scatterplots with log-transformation are closer together. Using log-transformations will improve the linear regression analysis since L_SHUCK and L_VOLUME result in a stronger linear relationship. Without log-transformations the plots are more clustered making it difficult to differentiate between the classes and the type. However, with log-transformations it is easier to identify class levels, especially A1 class which mostly in the bottom left corner of plot. A2 is closer to the center of the plot while A3, A4, and A5 are located more towards the top right corner. For TYPE, the infant data points are also more easily seen in the bottom right and the adults are located in the top right when using log-transformation.***


***####   Section 4: (5 points) ####***

(4)(a1) Since abalone growth slows after class A3, infants in classes A4 and A5 are considered mature and candidates for harvest. You are given code in (4)(a1) to reclassify the infants in classes A4 and A5 as ADULTS. 

```{r Part_4a1}

mydata$TYPE[mydata$CLASS == "A4" | mydata$CLASS == "A5"] <- "ADULT"
table(mydata$TYPE)

```

(4)(a2) Regress L_SHUCK as the dependent variable on L_VOLUME, CLASS and TYPE (Kabacoff Section 8.2.4, p. 178-186, the Data Analysis Video #2 and Black Section 14.2). Use the multiple regression model: L_SHUCK ~ L_VOLUME + CLASS + TYPE. Apply *summary()* to the model object to produce results.

```{r Part_4a2}
mydata$TYPE2 <- relevel(x = mydata$TYPE, ref ="I")
lm_mdl <- lm(L_SHUCK ~ L_VOLUME + CLASS + TYPE2, data = mydata)
summary(lm_mdl)

```

**Essay Question:  Interpret the trend in CLASS levelcoefficient estimates? (Hint:  this question is not asking if the estimates are statistically significant. It is asking for an interpretation of the pattern in these coefficients, and how this pattern relates to the earlier displays).**

***Answer: The trend in CLASS level coefficients show that as the CLASS level increases the coefficients get more negative resulting in a stronger negative correlation. This indicates that SHUCK weight decreases more as the abalones grow older and larger in volume. This pattern is consistent with the earlier displays.***

**Additional Essay Question:  Is TYPE an important predictor in this regression? (Hint:  This question is not asking if TYPE is statistically significant, but rather how it compares to the other independent variables in terms of its contribution to predictions of L_SHUCK for harvesting decisions.)  Explain your conclusion.**

***Answer: The TYPE variable does provide some value in a regression model for L_SHUCK with a .02 coefficient for adult abalones. However, compared to the other independent variables it is not as important since those coefficients are larger (except for the A2 class) meaning they are more important than TYPE as a predictor variable. ***

-----

The next two analysis steps involve an analysis of the residuals resulting from the regression model in (4)(a) (Kabacoff Section 8.2.4, p. 178-186, the Data Analysis Video #2).

-----

***#### Section 5: (5 points) ####***

(5)(a) If "model" is the regression object, use model$residuals and construct a histogram and QQ plot. Compute the skewness and kurtosis. Be aware that with 'rockchalk,' the kurtosis value has 3.0 subtracted from it which differs from the 'moments' package. 

```{r Part_5am, fig.width = 12}

par(mfrow = c(1, 2))
hist(lm_mdl$residuals, xlab = "Residuals", main = "Histogram of Linear Model Residuals")
qqnorm(lm_mdl$residuals, main = "QQ Plot for Linear Model Residuals")
qqline(lm_mdl$residuals)
```

(5)(b) Plot the residuals versus L_VOLUME, coloring the data points by CLASS and, a second time, coloring the data points by TYPE. Keep in mind the y-axis and x-axis may be disproportionate which will amplify the variability in the residuals. Present boxplots of the residuals differentiated by CLASS and TYPE (These four plots can be conveniently presented on one page using *par(mfrow..)* or *grid.arrange()*. Test the homogeneity of variance of the residuals across classes using *bartlett.test()* (Kabacoff Section 9.3.2, p. 222).  

```{r Part_5b, fig.width = 12, fig.height = 8}

grid.arrange(
  ggplot(data = mydata, aes(x = L_VOLUME, y = lm_mdl$residuals, col = CLASS)) + geom_point() + theme(legend.position="bottom"),
  ggplot(data = mydata, aes(x = L_VOLUME, y = lm_mdl$residuals, col = TYPE2)) + geom_point() + theme(legend.position="bottom"),
  ggplot(data = mydata, aes(x = CLASS, y = lm_mdl$residuals)) + geom_boxplot(),
  ggplot(data = mydata, aes(x = TYPE2, y = lm_mdl$residuals)) + geom_boxplot()
)

bartlett.test(x = lm_mdl$residuals, g = mydata$CLASS)

```

**Essay Question:  What is revealed by the displays and calculations in (5)(a) and (5)(b)? Does the model 'fit'?  Does this analysis indicate that L_VOLUME, and ultimately VOLUME, might be useful for harvesting decisions? Discuss.**  

***Answer: The displays from 5.a and 5.b show a normal distribution of the linear regression model residuals which appear to be centered around 0. Additionally, the p-value of 0.45 from the Bartlett test indicates that the variances of the residuals are relatively equal. This suggests that the linear regression model is a good fit and that abalone VOLUME is useful for harvesting decisions.  ***

-----

Harvest Strategy:

There is a tradeoff faced in managing abalone harvest. The infant population must be protected since it represents future harvests. On the other hand, the harvest should be designed to be efficient with a yield to justify the effort. This assignment will use VOLUME to form binary decision rules to guide harvesting. If VOLUME is below a "cutoff" (i.e. a specified volume), that individual will not be harvested. If above, it will be harvested. Different rules are possible.The Management needs to make a decision to implement 1 rule that meets the business goal.

The next steps in the assignment will require consideration of the proportions of infants and adults harvested at different cutoffs. For this, similar "for-loops" will be used to compute the harvest proportions. These loops must use the same values for the constants min.v and delta and use the same statement "for(k in 1:10000)."  Otherwise, the resulting infant and adult proportions cannot be directly compared and plotted as requested. Note the example code supplied below.

-----

***#### Section 6: (5 points) ####***

(6)(a) A series of volumes covering the range from minimum to maximum abalone volume will be used in a "for loop" to determine how the harvest proportions change as the "cutoff" changes. Code for doing this is provided.

```{r Part_6a}

idxi <- mydata$TYPE == "I"
idxa <- mydata$TYPE == "ADULT"

max.v <- max(mydata$VOLUME)
min.v <- min(mydata$VOLUME)
delta <- (max.v - min.v)/10000
prop.infants <- numeric(10000)
prop.adults <- numeric(10000)
volume.value <- numeric(10000)

total.infants <- sum(idxi)  
total.adults <- sum(idxa)

for (k in 1:10000) { 
	value <- min.v + k*delta
	volume.value[k] <- value
	prop.infants[k] <- sum(mydata$VOLUME[idxi] <= value)/total.infants
	prop.adults[k] <-  sum(mydata$VOLUME[idxa] <= value)/total.adults
}

```

(6)(b) Our first "rule" will be protection of all infants. We want to find a volume cutoff that protects all infants, but gives us the largest possible harvest of adults. We can achieve this by using the volume of the largest infant as our cutoff. You are given code below to identify the largest infant VOLUME and to return the proportion of adults harvested by using this cutoff. You will need to modify this latter code to return the proportion of infants harvested using this cutoff. Remember that we will harvest any individual with VOLUME greater than our cutoff.

```{r Part_6b}
# Largest infant volume
(max_inf_vol <- max(mydata$VOLUME[mydata$TYPE == "I"]))  # [1] 526.6383

# Proportion of adults harvested
sum(mydata$VOLUME[mydata$TYPE == "ADULT"] > max_inf_vol) /
  total.adults  # [1] 0.2476573

# Add code to calculate the proportion of infants harvested
sum(mydata$VOLUME[mydata$TYPE == "I"] > max_inf_vol) /
  total.infants

# If we use the largest infant volume, we harvest approximately 24.8% of adults and 0%,
# as expected, of infants.

```

(6)(c) Our next approaches will look at what happens when we use the median infant and adult harvest VOLUMEs. Using the median VOLUMEs as our cutoffs will give us (roughly) 50% harvests. We need to identify the median volumes and calculate the resulting infant and adult harvest proportions for both.

```{r Part_6c}
# Add code to determine the median infant volume:
(med_inf_vol <- median(mydata$VOLUME[mydata$TYPE == "I"]))

# Add code to calculate the proportion of infants harvested
sum(mydata$VOLUME[mydata$TYPE == "I"] > med_inf_vol) /
  total.infants

# Add code to calculate the proportion of adults harvested
sum(mydata$VOLUME[mydata$TYPE == "ADULT"] > med_inf_vol) /
  total.adults

# If we use the median infant volume as our cutoff, we harvest almost 50% of our infants
# and a little more than 93% of our adults.


# Add code to determine the median adult volume:
(med_adult_vol <- median(mydata$VOLUME[mydata$TYPE == "ADULT"]))

# Add code to calculate the proportion of infants harvested
sum(mydata$VOLUME[mydata$TYPE == "I"] > med_adult_vol) /
  total.infants

# Add code to calculate the proportion of adults harvested
sum(mydata$VOLUME[mydata$TYPE == "ADULT"] > med_adult_vol) /
  total.adults

# If we use the median adult volume as our cutoff, we harvest almost 50% of adults
# and approximately 2.4% of infants.

```

(6)(d) Next, we will create a plot showing the infant conserved proportions (i.e. "not harvested," the prop.infants vector) and the adult conserved proportions (i.e. prop.adults) as functions of volume.value. We will add vertical A-B lines and text annotations for the three (3) "rules" considered, thus far:  "protect all infants," "median infant" and "median adult." Your plot will have two (2) curves - one (1) representing infant and one (1) representing adult proportions as functions of volume.value - and three (3) A-B lines representing the cutoffs determined in (6)(b) and (6)(c).

```{r Part_6d, fig.width = 12, fig.height = 6}

colors <- c("Adults" = "blue", "Infants" = "red")

ggplot(data = NULL, aes(x = volume.value)) + geom_line(aes(y = prop.adults, color = "Adults"), size = 1.5) + 
  geom_line(aes(y = prop.infants, color = "Infants"), size = 1.5) + 
  scale_color_manual(values = colors) + 
  scale_y_continuous(name="Proportions", limits=c(0.00, 1.000)) + 
  labs(x= "Volume") + 
  theme(legend.title=element_blank()) + 
  geom_hline(yintercept = 0.5, linetype = "dashed") + 
  geom_vline(xintercept = med_inf_vol) + 
  geom_vline(xintercept = med_adult_vol) + 
  geom_vline(xintercept = max_inf_vol) + 
  annotate("text", x = 180, y = .47, label = "Median Infant") + 
  annotate("text", x = 157, y = .43, label = "133.8") + 
  annotate("text", x = 435, y = .47, label = "Median Adult") + 
  annotate("text", x = 412, y = .43, label = "384.6") + 
  annotate("text", x = 580, y = .97, label = "Largest Infant") + 
  annotate("text", x = 556, y = .93, label = "526.6") + 
  ggtitle("Proportions of Adults and Infants Protected") + 
  theme(plot.title = element_text(hjust = 0.5))
```

**Essay Question:  The two 50% "median" values serve a descriptive purpose illustrating the difference between the populations. What do these values suggest regarding possible cutoffs for harvesting?** 

***Answer: The median volume for the adult population is roughly 3 times larger than the infants which results in more infants being protected than adults if harvesting based on volume. These values suggest that volume can be a good measurement to use for possible cutoff thresholds for infant protection during abalone harvesting. ***

-----


More harvest strategies:

This part will address the determination of a cutoff volume.value corresponding to the observed maximum difference in harvest percentages of adults and infants. In other words, we want to find the volume value such that the vertical distance between the infant curve and the adult curve is maximum. To calculate this result, the vectors of proportions from item (6) must be used. These proportions must be converted from "not harvested" to "harvested" proportions by using (1 - prop.infants) for infants, and (1 - prop.adults) for adults. The reason the proportion for infants drops sooner than adults is that infants are maturing and becoming adults with larger volumes.

-----

***#### Section 7: (10 points)  ####***

(7)(a) Evaluate a plot of the difference ((1 - prop.adults) - (1 - prop.infants)) versus volume.value. Compare to the 50% "split" points determined in (6)(a). There is considerable variability present in the peak area of this plot. The observed "peak" difference may not be the best representation of the data. One solution is to smooth the data to determine a more representative estimate of the maximum difference.

```{r Part_7a}
diff <- ((1 - prop.adults) - (1 - prop.infants))
plot(y = diff, x = volume.value, type = "l")


```

(7)(b) Since curve smoothing is not studied in this course, code is supplied below. Execute the following code to create a smoothed curve to append to the plot in (a). The procedure is to individually smooth (1-prop.adults) and (1-prop.infants) before determining an estimate of the maximum difference. 

```{r Part_7b}

y.loess.a <- loess(1 - prop.adults ~ volume.value, span = 0.25,
	family = c("symmetric"))
y.loess.i <- loess(1 - prop.infants ~ volume.value, span = 0.25,
	family = c("symmetric"))
smooth.difference <- predict(y.loess.a) - predict(y.loess.i)

```

(7)(c) Present a plot of the difference ((1 - prop.adults) - (1 - prop.infants)) versus volume.value with the variable smooth.difference superimposed. Determine the volume.value corresponding to the maximum smoothed difference (Hint:  use *which.max()*). Show the estimated peak location corresponding to the cutoff determined.

Include, side-by-side, the plot from (6)(d) but with a fourth vertical A-B line added. That line should intercept the x-axis at the "max difference" volume determined from the smoothed curve here.

```{r Part_7c, fig.width = 12, fig.height = 6}
max_diff <- volume.value[which.max(smooth.difference)]

grid.arrange(
ggplot(data = NULL, aes( x= volume.value, y = diff)) + geom_line() +
  geom_line(aes(x = volume.value, y = smooth.difference), lty = 2, col = "red4") +
  geom_vline(xintercept = max_diff) + annotate("text", x = 280, y = 0.45, label = "Volume = 262.143", angle = 90) +
  xlab("Volume") + ylab("Difference in Proportions Harvested") + ggtitle("Difference Plot with Smoothed Curve") +
  theme(plot.title = element_text(hjust = 0.5)),
ggplot(data = NULL, aes(x = volume.value)) + geom_line(aes(y = prop.adults, color = "Adults"), size = 1.5) + 
  geom_line(aes(y = prop.infants, color = "Infants"), size = 1.5) + 
  scale_color_manual(values = colors) + scale_y_continuous(name="Proportions", limits=c(0.00, 1.000)) + 
  labs(x= "Volume") + theme(legend.title=element_blank()) + 
  geom_hline(yintercept = 0.5, linetype = "dashed") + 
  geom_vline(xintercept = med_inf_vol) + 
  geom_vline(xintercept = med_adult_vol) + 
  geom_vline(xintercept = max_inf_vol) + 
  annotate("text", x = 160, y = .35, label = "Median Infant = 133.8", angle = 90) + 
  annotate("text", x = 405, y = .35, label = "Median Adult = 384.6", angle = 90) + 
  annotate("text", x = 555, y = .35, label = "Largest Infant = 526.6", angle = 90) + 
  ggtitle("Proportions of Adults and Infants Protected") + 
  theme(plot.title = element_text(hjust = 0.5)) + 
  geom_vline(xintercept = max_diff) + 
  annotate("text", x = 290, y = .7, label = "Max Diff = 262.1", angle = 90),
ncol = 2
)
```

(7)(d) What separate harvest proportions for infants and adults would result if this cutoff is used? Show the separate harvest proportions. We will actually calculate these proportions in two ways:  first, by 'indexing' and returning the appropriate element of the (1 - prop.adults) and (1 - prop.infants) vectors, and second, by simply counting the number of adults and infants with VOLUME greater than the vlume threshold of interest.

Code for calculating the adult harvest proportion using both approaches is provided.

```{r Part_7d}

(1 - prop.adults)[which.max(smooth.difference)]  # [1] 0.7416332
# OR,
sum(mydata[mydata$TYPE == "ADULT", "VOLUME"] >
      volume.value[which.max(smooth.difference)]) / total.adults # [1] 0.7416332

(1 - prop.infants)[which.max(smooth.difference)]

```

-----

There are alternative ways to determine cutoffs. Two such cutoffs are described below.

-----

***####  Section 8: (10 points)  ####***

(8)(a) Harvesting of infants in CLASS "A1" must be minimized. The smallest volume.value cutoff that produces a zero harvest of infants from CLASS "A1" may be used as a baseline for comparison with larger cutoffs. Any smaller cutoff would result in harvesting infants from CLASS "A1."  

Compute this cutoff, and the proportions of infants and adults with VOLUME exceeding this cutoff. Code for determining this cutoff is provided. Show these proportions. You may use either the 'indexing' or 'count' approach, or both.

```{r Part_8a}

volume.value[volume.value > max(mydata[mydata$CLASS == "A1" &
  mydata$TYPE == "I", "VOLUME"])][1] # [1] 206.786

sum(mydata[mydata$TYPE == "I", "VOLUME"] >
  volume.value[volume.value > max(mydata[mydata$CLASS == "A1" &
  mydata$TYPE == "I", "VOLUME"])][1]) / total.infants 

sum(mydata[mydata$TYPE == "ADULT", "VOLUME"] >
  volume.value[volume.value > max(mydata[mydata$CLASS == "A1" &
  mydata$TYPE == "I", "VOLUME"])][1]) / total.adults 


```

(8)(b) Next, append one (1) more vertical A-B line to our (6)(d) graph. This time, showing the "zero A1 infants" cutoff from (8)(a). This graph should now have five (5) A-B lines:  "protect all infants," "median infant," "median adult," "max difference" and "zero A1 infants."

```{r Part_8b, fig.width = 12}

ggplot(data = NULL, aes(x = volume.value)) + geom_line(aes(y = prop.adults, color = "Adults"), size = 1.5) + 
  geom_line(aes(y = prop.infants, color = "Infants"), size = 1.5) + 
  scale_color_manual(values = colors) + 
  scale_y_continuous(name="Proportions", limits=c(0.00, 1.000)) + labs(x= "Volume") + 
  theme(legend.title=element_blank()) + 
  geom_hline(yintercept = 0.5, linetype = "dashed") + 
  geom_vline(xintercept = med_inf_vol) + 
  geom_vline(xintercept = med_adult_vol) + 
  geom_vline(xintercept = max_inf_vol) + 
  annotate("text", x = 145, y = .35, label = "Median Infant = 133.8", angle = 90) + 
  annotate("text", x = 400, y = .35, label = "Median Adult = 384.6", angle = 90) + 
  annotate("text", x = 537, y = .25, label = "Largest Infant = 526.6 (protect all infants)", angle = 90) + 
  ggtitle("Proportions of Adults and Infants Protected") + theme(plot.title = element_text(hjust = 0.5)) +
  geom_vline(xintercept = max_diff) + 
  annotate("text", x = 275, y = .7, label = "Max Diff = 262.1", angle = 90) + 
  geom_vline(xintercept = 206.786) + 
  annotate("text", x = 215, y = .35, label = "zero A1 infants = 262.1", angle = 90)

```


***#### Section 9: (5 points) ####***

(9)(a) Construct an ROC curve by plotting (1 - prop.adults) versus (1 - prop.infants). Each point which appears corresponds to a particular volume.value. Show the location of the cutoffs determined in (6), (7) and (8) on this plot and label each. 

```{r Part_9, fig.width = 8.5}
ggplot(data = NULL, aes(x = (1 - prop.infants), y = (1 - prop.adults))) + geom_line(size = 1) + 
  geom_abline(slope = 1, intercept = 0, linetype = 2, color = "red4") +  
  geom_point(aes(x=0, y=0.2476573), colour="blue", size = 5, shape = 1, stroke = 1) + 
  annotate("text", x= .09, y = .2, label = "Protect all infants \n vol = 526.6") + 
  geom_point(aes(x= 0.4982699, y= 0.9330656), colour="blue", size = 5, shape = 1, stroke = 1) +
  annotate("text", x= .5, y = .85, label = "Median infants \n vol = 133.8") +
  geom_point(aes(x= 0.02422145, y= 0.4993307), colour="blue", size = 5, shape = 1, stroke = 1) +
  annotate("text", x= .1, y = .45, label = "Median adults \n vol = 384.6") +
  geom_point(aes(x = 0.1764706, y= 0.7416332), colour="blue", size = 5, shape = 1, stroke = 1) +
  annotate("text", x= .23, y = .68, label = "Max difference \n vol = 262.1") +
  geom_point(aes(x =  0.2871972, y= 0.8259705), colour="blue", size = 5, shape = 1, stroke = 1) +
  annotate("text", x= .26, y = .90, label = "Zero A1 infants \n vol = 206.8")
                                                                                                                            

```

(9)(b) Numerically integrate the area under the ROC curve and report your result. This is most easily done with the *auc()* function from the "flux" package.   Areas-under-curve, or AUCs, greater than 0.8 are taken to indicate good discrimination potential. 

```{r Part_9b}
auc(x = (1-prop.infants), y = (1-prop.adults))

```


***#### Section 10: (10 points) ####***

(10)(a) Prepare a table showing each cutoff along with the following:
 	1) true positive rate (1-prop.adults,
 	2) false positive rate (1-prop.infants),
 	3) harvest proportion of the total population

To calculate the total harvest proportions, you can use the 'count' approach, but ignoring TYPE; simply count the number of individuals (i.e. rows) with VOLUME greater than a given threshold and divide by the total number of individuals in our dataset.
 	
```{r Part_10} 	

#protect all
sum(mydata$VOLUME > max_inf_vol) / nrow(mydata)

#median infants
sum(mydata$VOLUME > med_inf_vol) / nrow(mydata)

#median adult
sum(mydata$VOLUME > med_adult_vol) / nrow(mydata)

#max diff
sum(mydata$VOLUME >
      volume.value[which.max(smooth.difference)]) / nrow(mydata)

#zero A1 infants
sum(mydata$VOLUME >
  volume.value[volume.value > max(mydata[mydata$CLASS == "A1" &
  mydata$TYPE == "I", "VOLUME"])][1]) / nrow(mydata) 

rnames <- c("Protect all Infants", "Median Infants", "Median Adults", "Max Difference", "Zero A1 Infants")
vols <- c(526.6, 133.8, 384.6, 262.1, 206.8)
TPR <- c(0.248, 0.933, 0.499, 0.742, 0.826)
FPR <- c(0.000, 0.498, 0.024, 0.176, 0.287)
total_harvest <- c(0.179, 0.812, 0.367, 0.584, 0.676)

tbl <- cbind(rnames, vols, TPR, FPR, total_harvest)
colnames(tbl) <- c("Cutoff Description", "Volume", "TPR", "FPR", "Total Harvest")
kable(tbl)

```
 	
**Essay Question: Based on the ROC curve, it is evident a wide range of possible "cutoffs" exist. Compare and discuss the five cutoffs determined in this assignment.**   

***Answer: The Protect all Infants cutoff results in the highest volume threshold but the lowest TPR and lowest proportion of abalones harvested. The median infant cutoff results in the lowest volume threshold which leads to it being the highest proportion of abalones harvested. This cutoff also produces the highest TPR and the highest FPR. The median adult cutoff produces the lowest FPR and second lowest proportion of abalones harvested. The maximum difference cutoff produces the median TPR and median FPR which results in a little over half of the abalones being harvested. Lastly, the Zero A1 Infant cutoff has the second highest TPR and FPR among the different thresholds. This results in about 2/3 of the abalones being harvest which is also the second highest.***



**Final Essay Question:  Assume you are expected to make a presentation of your analysis to the investigators How would you do so?  Consider the following in your answer:**

1. Would you make a specific recommendation or outline various choices and tradeoffs?
2. What qualifications or limitations would you present regarding your analysis?
3. If it is necessary to proceed based on the current analysis, what suggestions would you have for  implementation of a cutoff?  
4. What suggestions would you have for planning future abalone studies of this type? 

***Answer:***   
***1. I would not make a specific recommendation on determining the cutoff threshold for abalone harvesting. However, I would describe in detail the various choices for cutoff thresholds and the tradeoffs of each choice based on the analysis above. This is because depending on the desired results, different cutoffs may be better suited for different situations. For example, if the abalone population is over-populated, then the median infant cutoff might be appropriate because it results in the highest proportion harvested. On the other hand if the abalone population is under-population and endangered, then the Protect all infants might be more appropriate in this case.***  
  
***2. The qualifications or limitations I would present regarding this analysis is that this study used observational data in which there was no control in selecting the sample population. Additionally, I do not know what steps were taken if any to avoid any sampling bias. I would also mention that the were outliers present in the data which can be seen in the box plots.***  
  
***3. If it is necessary to process based on the current analysis, I would suggest to implement a conservative approach in order to protect infants but still be able to harvest a large enough portion to meet consumer demand. For this reason, I would most likely suggest either the Max Difference cutoff or Zero A1 Infants cutoff.***  
  
***4. In terms of planning future abalone studies of this type, I would suggest increasing the sample size and also analyzing additional predictive variables to see if there are other factors that could potentially improve the model by decreasing the amount of false positives. If multiple variables can be used to explain the variability, then perhaps a multiple linear regression model can be used instead. Lastly, I would suggest taking the appropriate steps in order to minimize as much sampling bias as possible.***